Estimation of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2013-12-26 paper/tsunami.tex
Some tsunmai changes

\section{Manning's Friction Coefficient}
%We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.

Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
water equations governing tsunamis and the roughness parameter $n$. 
This Manning's $n$ coefficient is variable that is highly spatial 
and dependent on the surface characteristics of the sea-floor. 
The direct measurement of $n$ is therefore impossible, instead, values
are often assigned from tables that are empirically estimated
for common types of land that are determined from satellite images.
As a result, the Manning's $n$ coefficients used in tsunami models
often contain large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties
in the predicted quantities of interest such as the water elevations
and becomes a health risk during tsunami events. In addition, 
the prediction of these quantities of interest are often highly sensitive to 
the changes in these uncertain parameters. 
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $n$ by three different parameters
that are varied in three different regions but are assumed
constant in each region. The three regions are as follows: 
on-shore whose coefficient is denoted by $n_1$, near-shore 
denoted by $n_2$, and deep-water denoted by $n_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform 
distribution was assumed as the prior where $n$ was sampled in the interval $n \in [0.005,0.2]$.

@@

The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:

\begin{equation} \label{eq:swe}
    \begin{aligned}
    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
    \end{aligned}
\end{equation}

where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
longitudinal and latitudinal directions respectively, $g$ the acceleration due 
to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
given a Manning's $n$ parameterization such that:

\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\end{equation}
where $n$ is an empirically determined parameter.

The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.  The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.

One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, a tsunami wave).  \geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.

\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
at the gauges ready.... Not sure if this explanation is worth including here ///}

@@

Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.  After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and

\begin{itemize}
    \item Perhaps do one of the tsunami inversion fault models
\end{itemize}

@@

\subsubsection{Bathymetry}
The bathymetry is shown in Figure~\ref{fig:setup} in addition to the topography
of the Pacific Ocean.
\subsubsection{Observations}

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).

>>>>

%!TEX root = paper.tex

@@

The \geoclaw software package uses a finite volume, wave-propagation approach described in \cite{LeVeque:1997eg} to solve the two-dimensional shallow water equations:
\begin{equation} \label{eq:swe}
    \begin{aligned}
    &\pd{}{t} h + \pd{}{x} (hu) + \pd{}{y} (hv) = 0, \\
    &\pd{}{t}(hu) + \pd{}{x} \left(hu^2 + \frac{1}{2} g h^2 \right ) + \pd{}{y} (huv) = ~~ fhv - gh \pd{}{x} b - C_f |\vec{u}| hu \\
    &\pd{}{t} (hv) + \pd{}{x} (huv) + \pd{}{y} \left (hv^2 + \frac{1}{2} gh^2 \right) = -fhu - gh \pd{}{y} b - C_f |\vec{u}| hv
    \end{aligned}
\end{equation}
where $h$ is the depth of the water column, $u$ and $v$ the velocities in the 
longitudinal and latitudinal directions respectively, $g$ the acceleration due 
to gravity, $b$ the bathymetry, $f$ the Coriolis parameter, and $C_f$ the bottom 
friction coefficient.  As is common in tsunami modeling, $C_f$ is calculated 
given a Manning's $n$ parameterization such that:
\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\end{equation}
where $n$ is an empirically determined parameter known as Manning's $n$.

The primary computational kernel in \geoclaw is the evaluation of the solution to the Riemann problem at each grid cell interface.  The Riemann solver used includes the ability to handle inundation (wet-dry interfaces), well-balancing, even when the momentum is non-zero, and entropy corrections \cite{George:2008aa}.

\subsubsection{Adaptivity} \label{ssub:adaptivity}
One unique feature that \geoclaw has is the ability to adaptively refine the grids used for the computation essentially following a region of interest as time progresses (in this case, the disturbance in the surface height $\eta$).  \geoclaw implements these algorithms following \cite{Berger:1984ui,Berger:1998aa} and is described in detail in \cite{Berger:2011du} in the case of tsunami modeling.

For the simulations presented, refinement thresholds were used that matched what was presented in \cite{MacInnes:2013cr}.  As was done there, four levels of refinement were used staring with a resolution of 1 degree in both the latitudinal and longitudinal directions down to 0.5' resolution (approximately 900 meters) located around the observation locations.  The tolerance for the refinement criteria for sea-surface anomaly was 0.02 meters.

@@

The MacInnes \emph{et al.} study found that source models from tsunami
inversions were

Since the previous study \cite{MacInnes:2013cr} did not find that the source model effected the tsunami wave-forms at the DART buoys, we did not expect there to be much difference between the available models.  After comparing the same analysis with two different source models, namely the models from \cite{Ammon:2011dm} and .

[4] Ammon used a seismic and GPS inversion technique
[7] Saito et al. used a tsunami inversion

@@

\subsubsection{Bathymetry}
The bathymetry used in the simulation is a combination of ETOPO 1' and 4' 
accurate bathymetry for the region \cite{Amante:2009ud}.  The additional 
bathymetry data used in the MacInnes \emph{et al.} was useful only near the
shore and had a smaller effect on the regions near the observations used in
this study.

\subsubsection{Observations}
The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).

\section{Manning’s $n$ Friction Parameterization}
%We use the widely used .  The parameter $n$ is also allowed to vary in space determined by the initial water depth to study effects of deep-water, near-shore, and on-shore sensitivities.

Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
water equations governing tsunamis and the roughness parameter $n$. 
This Manning's $n$ coefficient is variable that is highly spatial 
and dependent on the surface characteristics of the sea-floor. 
The direct measurement of $n$ is therefore impossible, instead, values
are often assigned from tables that are empirically estimated
for common types of land that are determined from satellite images.
As a result, the Manning’s n coefficients used in tsunami models
often contain large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties
in the predicted quantities of interest such as the water elevations
and becomes a health risk during tsunami events. In addition, 
the prediction of these quantities of interest are often highly sensitive to 
the changes in these uncertain parameters. 
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $n$ by three different parameters
that are varied in three different regions but are assumed
constant in each region. The three regions are as follows: 
on-shore whose coefficient is denoted by $n_1$, near-shore 
denoted by $n_2$, and deep-water denoted by $n_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform distribution was assumed as the prior where $N$ was sampled in the interval $n \in [0.005-0.2]$.

Note that on the Sendai plain a Manning's $n$ value of 0.035 was used to account for pasture, farmland and rice paddies.
